Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-03-14 documents/ms.tex
last few edits

Standard non-inference based methods to measure rotation periods include
detecting peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS)
periodogram \citep[\eg][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep[\eg][]{Mcquillan13b} and wavelet transforms \citep[\eg][]{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram, and a
choice of mother wavelet, assumed to describe the data over a range of
transpositions \citep[see, \eg][]{Carter2010}, for the wavelet method.
In contrast, since it does not rely on a generative model, the ACF method is
much better suited to signals that are non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such as
the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.

@@

GPs are commonly used in the machine learning community and increasingly
in other scientific fields such as biology, geophysics and cosmology.
More recently, GPs have been used in the stellar and exoplanet fields within
astronomy, to capture stellar variability or instrumental systematics
\citep[see \eg][]{Gibson2012, Haywood2014, Dawson2014, Barclay2015,
Haywood2015, Evans2015, Rajpaul2015, Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset.
An in-depth introduction to Gaussian processes is provided in
\citet{Rasmussen2005}.

@@

A range of kernel functions can describe stellar variability.
For example, the most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, could adequately fit the
KIC \kepexample\ light curve:
\begin{equation}
\label{eq:SE}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
Here $A>0$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}
However, The SE kernel function does not well describe the covariance
in stellar light curves, nor is it {\it useful} for the problem of
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel.
\citet{Rasmussen2005} model QP variability in CO$_2$ concentration on the
summit of the Mauna Loa volcano in Hawaii \citep[data from][]{Keeling2004}
using a kernel which is the product of a periodic and a SE kernel: the QP
kernel.
This kernel is defined as
\begin{equation}
\label{eq:QP}
k_{i,j} = A \exp \left[-\frac{(x_i - x_j)^2}{2l^2} -
    \Gamma^2 \sin^2\left(\frac{\pi(x_i - x_j)}{P}\right) \right] + \sigma^2
    \delta_{ij}.
\end{equation}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very large, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is small, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
In other words, large values of $\Gamma$ lead to periodic variations with
increasingly complex harmonic content.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing
points separated by half a period to be weakly correlated.
We also use an extra parameter, $\sigma$, which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This can be interpreted to represent underestimation of observational
uncertainties~---~if the uncertainties reported on the data are too small, it
will be non-zero~---~or it can capture any remaining ``jitter'' or residuals
that are not captured by the effective GP model.
We use this QP kernel function to produce the GP model
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.
There are many ways to construct a QP kernel function, involving a range of
choices for both the periodic and a-periodic components of the model.
The kernel function presented above reproduces the behaviour of stellar light
curves but it is not necessarily the only model choice that can do so.
We do not attempt to test any other choices in this paper, noting only that
this kernel function provides an adequate fit to the data.
We leave formal comparisons with other kernel function choices to a future
publication.

@@

The GP likelihood is similar to the simple Gaussian likelihood
function used for optimisation problems where the uncertainties are
Gaussian and uncorrelated. The latter can be written
\begin{equation}
    \ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\left[\frac{(y_n-\mu)^2}{\sigma_n^2}
    + \ln(2\pi\sigma_n^2)\right],
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainties on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \frac{N}{2}\log2\pi,
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
    \mathbf{C} &=& \left (\begin{array}{cccc}
    \sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
    \sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
    \sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!  Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function
${\bf K}$ replaces ${\bf C}$ in the above equation (for our purposes, the QP
kernel of Equation \ref{eq:QP}).

Evaluating this likelihood for a large number of points can be computationally
expensive.  For example, evaluating
$\mathcal L$ for an entire \Kepler\ lightcurve
($\sim$40,000 points) takes about $\sim$5\,s--- too slow
to perform inference on large numbers of light curves\footnote{All
computational times cited in this section are based on evaluations on a
single core of a 2015 Macbook Pro, 3.1 GHz Intel Core i7.}.
The matrix operations necessary to evaluate the GP likelihood\footnote{These
operations use the fast matrix solver HODLR \citep{Ambikasaran2014},
implemented in the {\tt george} \citep{George} python package.} scale as
$N\ln(N)^2$, where $N$ is the number of data points in the light curve.

@@

The flexibility of this GP model allows for posterior multimodality and
``over-fitting''--like behavior.
For example, if $l$ is small, the non-periodic factor in the covariance
kernel may dominate, allowing for a good fit to the data without
requiring any periodic covariance structure---even if clear periodic
structure is present.
Additionally, for large values of $\Gamma$ the GP model becomes extremely
flexible and can fit the data without varying the period.
Managing this flexibility to reliably retrieve the period parameter requires
imposing informative priors on the non-period GP parameters.
% While there are no strictly {\it a priori} reasons to believe
% that any of the hyperparameters should take values in any specific range,
% we have found in this work that reasonable priors for any particular
% dataset can be determined by experimentation.
% We note this is a somewhat subjective art, with a goal of
% not allowing the GP to become so flexible that it does not require
% periodicity to fit the data even in the presence of clear periodicity.
In particular, we find it necessary to avoid large values of
$A$ and $\Gamma$, and small values of $l$; though the exact details of what
works well may differ among datasets.
See \textsection \ref{sec:discussion} for a discussion of more physically
motivated priors on the non-period parameters.

@@

In order to benchmark this new rotation period recovery method, we apply
it to a set of simulated light curves
and compare to the performance of established literature methods.
Section \ref{sec:simulations} describes the simulated data we use;
Section \ref{sec:performance} demonstrates the performance of the
QP-GP method; and Section \ref{sec:comparison} compares it to the performance
of the Lomb-Scargle periodogram and autocorrelation function methods.

@@

We apply the QP-GP inference method described in Section \ref{sec:method}
to each of these 333 simulated light curves.  As discussed in Section
\ref{sec:nonperiod_prior}, reliable inference
requires defining a useful set of priors on the non-period
hyperparameters.
For this simulation dataset, we determined these by first running
the method using very broad priors on all the non-period parameters
(log-flat between -20 and 20)
and then inspecting the distribution of their posteriors for those cases
that successfully recovered the true period.
We also experimented with constraining the allowed ranges of the parameters
after discovering that some regions of parameter space (such as large values
of $A$ and $\Gamma$ and small values of $l$) tended to allow fits that ignored
the desired periodicity.
We list the final priors and bounds this process led us to adopt
in Table \ref{tab:priors}.
These priors were determined experimentally --- see \textsection
\ref{sec:discussion} for a discussion on more physically motivated priors.
% We note that we do not have any specific quantitative justification for the
% use of these priors, except that they produce good results for this particular
% test set of light curves.
% We thus caution that other data sets might have different noise properties,
% potentially requiring modifications to these priors \vrcomment{[VR: might
% want to reword or remove these remarks; see my earlier comments]}
% (e.g., see Section \ref{sec:kepler}).
For the period prior, we tried two different methods: an uninformed (log-flat)
prior between 0.5\,d and 100\,d, and an ACF-informed prior
(Section \ref{sec:period_prior}).

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarise our results compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
To assess the performance of the QP-GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters, for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.
% The posterior samples of the model parameters for the simulated light curves
% are available online:
% \url{https://zenodo.org/record/292340\#.WKWpiBIrJE4}.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_acfprior_03_10.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in \textsection \ref{sec:GP_prior}.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

@@

The ACF method has proven extremely useful for measuring rotation periods.
The catalogue of rotation periods of \Kepler\ stars provided in
\citet{Mcquillan2013} has been widely used by the community and has provided
ground-breaking results for stellar and exoplanetary science.
The method also performed well in the \citet{Aigrain2015}
recovery experiment,
producing a large number of accurate rotation period measurements
(see, e.g., their Figure 8).
Another advantage is its fast implementation speed.
However, because the ACF method is non-probabilistic, ACF-estimated
rotation period uncertainties are poorly defined---a clear
disadvantage.
Figure \ref{fig:demo_lc} contains an example ACF of the light curve in
Figure \ref{fig:demo_acf}.

We compare our results to those of the Tel Aviv team in \citet{Aigrain2015}
who use the ACF method described in \citet{Mcquillan2013}.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The MRD of their periods, relative to the true rotation periods, is
\percenttelavivMAD\% --- smaller than the results produced by the GP
method.
However, their results contain a larger number of extreme outliers --- they
predict
periods that are significantly different from the true period in more cases
than the GP method.
This is revealed by comparing the root-mean-squared (RMS) of the two sets of
results since the RMS is more sensitive to outliers.
The RMS of the Tel Aviv method results is \telavivRMS\ days and the RMS of GP
method results is \gpRMS\ days.
The MRD of the ACF-recovered periods is \percentacfMAD\% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The injected and recovered rotation periods generally agree well,
though with a few drastic over- or underestimates.
Additionally, of the points clustered close to the 1:1 line,
more fall slightly below than above it, \ie\
rotation periods are systematically underestimated.
This stems from peak position measurements in the ACF method.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left (towards
shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice
simply measures the peak position without taking it into account.

@@

The \citet{Aigrain2015} Tel Aviv team were the only group to report periods
for all \naigrain\ stars, and are therefore the only team with which we
compare results from this paper.
The other participating groups chose to report only the period measurements
they were confident in and are likely to have omitted large outliers.
So, in order to test our results against an additional commonly used
technique, we implemented a simple LS periodogram method.

For each simulated light curve, we compute a LS periodogram\footnote{LS
periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.} over a
grid of 10,000 periods, evenly spaced in frequency, between 1 and 100 days.
We adopt the period of the highest peak in the periodogram as the rotation
period.
Figure \ref{fig:pgram_compare} shows the resulting recovered rotation periods
as a function of true period.
The MRD of the periodogram-recovered periods is \percentpgramMAD\% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
This method drastically overestimates many rotation periods,
because long-term trends unrelated to rotation can cause excess power
at long periods.
% a problem further exacerbated by instrumental noise.

@@

\begin{table*}
\begin{center}
    \caption{Median absolute (MAD), median relative (MRD) deviations and
    root-mean-squared (RMS) for
    the LS periodogram, ACF, \citet{Mcquillan2013} ACF and GP period recovery
    methods.}
\begin{tabular}{lcccc}
    Method & MAD & MRD & RMS \\
    \hline
    LS periodogram & \pgramMAD\ days & \percentpgramMAD \% & \pgramRMS\ days\\
    ACF & \telavivMAD\ days & \percenttelavivMAD \%
    & \telavivRMS\ days \\
    GP (uninformative prior) & \gpMADnp\ days & \percentgpMADnp \% &
    \gpRMSnp\ days \\
    GP (acf prior) & \gpMAD\ days & \percentgpMAD \% & \gpRMS\ days \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:MADs}

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, with only a few
exceptions, demonstrating that this method works not only on simulated data,
but also on real data---with the caveat that for any particular data set, some
care is needed regarding the setting the priors and ranges for the GP
hyperparameters.
Notably, the GP method recovers periods
systematically slightly larger than \citet{Mcquillan2013}---a likely consequence
of correcting for the ACF peak measurement bias discussed in Section \ref{sec:acf}.

>>>>

Standard non-inference based methods to measure rotation periods include
detecting peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS)
periodogram \citep[\eg][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep[\eg][]{Mcquillan13b} and wavelet transforms \citep[\eg][]{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram, and a
choice of mother wavelet, assumed to describe the data over a range of
transpositions and scales \citep[see, \eg][]{Carter2010}, for the wavelet
method.
In contrast, since it does not rely on a generative model, the ACF method is
much better suited to signals that are non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such as
the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.

@@

GPs are commonly used in the machine learning community and increasingly
in other scientific fields such as biology, geophysics and cosmology.
More recently, GPs have been used in the stellar and exoplanet fields within
astronomy, to capture stellar variability or instrumental systematics
\citep[see \eg][]{Gibson2012, Haywood2014, Dawson2014, Barclay2015,
Haywood2015, Evans2015, Rajpaul2015, Czekala2015, Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset.
An in-depth introduction to Gaussian processes is provided in
\citet{Rasmussen2005}.

@@

A range of covariance models, or kernel functions, can describe stellar
variability.
For example, the most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, could adequately fit the
KIC \kepexample\ light curve:
\begin{equation}
\label{eq:SE}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
Here $A>0$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}
However, The SE kernel function does not well describe the covariance
in stellar light curves, nor is it {\it useful} for the problem of
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel.
\citet{Rasmussen2005} model QP variability in CO$_2$ concentration on the
summit of the Mauna Loa volcano in Hawaii \citep[data from][]{Keeling2004}
using a kernel which is the product of a periodic and a SE kernel: the QP
kernel.
This kernel is defined as
\begin{equation}
\label{eq:QP}
k_{i,j} = A \exp \left[-\frac{(x_i - x_j)^2}{2l^2} -
    \Gamma^2 \sin^2\left(\frac{\pi(x_i - x_j)}{P}\right) \right] + \sigma^2
    \delta_{ij}.
\end{equation}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very large, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is small, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
In other words, large values of $\Gamma$ lead to periodic variations with
increasingly complex harmonic content.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing
points separated by half a period to be weakly correlated.
We also use an extra parameter, $\sigma$, which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This can be interpreted to represent underestimation of observational
uncertainties~---~if the uncertainties reported on the data are too small, it
will be non-zero~---~or it can capture any remaining ``jitter'' or residuals
that are not captured by the effective GP model.

We use this QP kernel function, defined in equation \ref{eqn:QP} to produce
the GP model that fits the \Kepler\ light curve in in Figure
\ref{fig:GP_example}.
There are many ways to construct a QP kernel function, involving a range of
choices for both the periodic and a-periodic components of the model.
The kernel function presented above reproduces the behaviour of stellar light
curves but it is not necessarily the only model choice that can do so.
We do not attempt to test any other choices in this paper, noting only that
this kernel function provides an adequate fit to the data.
We leave formal comparisons with other kernel function choices to a future
publication.

@@

The GP likelihood is similar to the simple Gaussian likelihood
function used for optimisation problems where the uncertainties are
Gaussian and uncorrelated. The latter can be written
\begin{equation}
    \ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\left[\frac{(y_n-\mu)^2}{\sigma_n^2}
    + \ln(2\pi\sigma_n^2)\right],
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainties on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \frac{N}{2}\log2\pi,
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
    \mathbf{C} &=& \left (\begin{array}{cccc}
    \sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
    \sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
    \sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!  Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function
${\bf K}$ replaces ${\bf C}$ in the above equation (for our purposes, the QP
kernel of Equation \ref{eq:QP} generates ${\bf K}$).

Evaluating this likelihood for a large number of points can be computationally
expensive.
For example, evaluating $\mathcal L$ for an entire \Kepler\ lightcurve
($\sim$40,000 points) takes about $\sim$5\,s--- too slow to perform inference
on large numbers of light curves\footnote{All computational times cited in
this section are based on evaluations on a single core of a 2015 Macbook Pro,
3.1 GHz Intel Core i7.}.
The matrix operations necessary to evaluate the GP likelihood scale as
$N\ln(N)^2$, where $N$ is the number of data points in the light curve.
These operations use the fast matrix solver HODLR \citep{Ambikasaran2014},
implemented in the {\tt george} \citep{George} python package.

@@

The flexibility of this GP model allows for posterior multimodality and
``over-fitting''--like behavior.
For example, if $l$ is small, the non-periodic factor in the covariance
kernel may dominate, allowing for a good fit to the data without
requiring any periodic covariance structure---even if clear periodic
structure is present.
Additionally, for large values of $\Gamma$ the GP model becomes extremely
flexible and can fit the data without varying the period.
Managing this flexibility to reliably retrieve the period parameter requires
imposing informative priors on the non-period GP parameters.
% While there are no strictly {\it a priori} reasons to believe
% that any of the hyperparameters should take values in any specific range,
% we have found in this work that reasonable priors for any particular
% dataset can be determined by experimentation.
% We note this is a somewhat subjective art, with a goal of
% not allowing the GP to become so flexible that it does not require
% periodicity to fit the data even in the presence of clear periodicity.
In particular, we find it necessary to avoid large values of
$A$ and $\Gamma$, and small values of $l$; though the exact details of what
works well may differ among datasets.
% See \textsection \ref{sec:discussion} for a discussion of more physically
% motivated priors on the non-period parameters.

@@

In order to test this new rotation period recovery method, we apply it to a
set of simulated light curves and compare to the performance of established
literature methods.
Section \ref{sec:simulations} describes the simulated data we use;
Section \ref{sec:performance} demonstrates the performance of the
QP-GP method; and Section \ref{sec:comparison} compares it to the performance
of the Lomb-Scargle periodogram and autocorrelation function methods.

@@

We apply the QP-GP inference method described in Section \ref{sec:method}
to each of these 333 simulated light curves.  As discussed in Section
\ref{sec:nonperiod_prior}, reliable inference
requires defining a useful set of priors on the non-period
hyperparameters.
For this simulation dataset, we determined these by first running
the method using very broad priors on all the non-period parameters
(log-flat between -20 and 20)
and then inspecting the distribution of their posteriors for those cases
that successfully recovered the true period.
We also experimented with constraining the allowed ranges of the parameters
after discovering that some regions of parameter space (such as large values
of $A$ and $\Gamma$ and small values of $l$) tended to allow fits that ignored
the desired periodicity.
We list the final priors and bounds this process led us to adopt
in Table \ref{tab:priors}.
% These priors were determined experimentally --- see \textsection
% \ref{sec:discussion} for a discussion on more physically motivated priors.
% We note that we do not have any specific quantitative justification for the
% use of these priors, except that they produce good results for this particular
% test set of light curves.
% We thus caution that other data sets might have different noise properties,
% potentially requiring modifications to these priors \vrcomment{[VR: might
% want to reword or remove these remarks; see my earlier comments]}
% (e.g., see Section \ref{sec:kepler}).
For the period prior, we tried two different methods: an uninformed (log-flat)
prior between 0.5\,d and 100\,d, and an ACF-informed prior
(Section \ref{sec:period_prior}).

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarise our results compared to the injected `true' stellar rotation
periods, for the uninformed and ACF-informed priors on period, respectively.
The periods inferred using the GP method show good agreement with the true
periods.
To assess the performance of the QP-GP and other period recovery methods, we
compute Median Absolute Deviations (MADs) of the results, relative to the
input periods.
We also compute the Median Relative Deviation (MRD), as a percentage and the
root-mean-squared (RMS).
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters, for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.
% The posterior samples of the model parameters for the simulated light curves
% are available online:
% \url{https://zenodo.org/record/292340\#.WKWpiBIrJE4}.

@@

\begin{figure*}
\begin{center}
% \includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\includegraphics[width=6in, clip=true]{figures/comparison_acfprior_03_10.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the GP
technique with an informed, ACF-based prior.
    Points are coloured by the peak-to-peak amplitude of the light curve, as
    defined in \citet{Aigrain2015}.
Since the posterior PDFs of rotation periods are often non-Gaussian,
    the points plotted here are maximum {\it a-posteriori} results.
The uncertainties are the 16th and 84th percentiles.
In many cases, the uncertainties are under-estimated.
The ACF-informed prior on rotation period used to generate these results is
    described in \textsection \ref{sec:GP_prior}.
    }
\label{fig:compare_mcmc_acfprior}
\end{center}
\end{figure*}

@@

The ACF method has proven extremely useful for measuring rotation periods.
The catalogue of rotation periods of \Kepler\ stars provided in
\citet{Mcquillan2013} has been widely used by the community and has provided
ground-breaking results for stellar and exoplanetary science.
The method also performed well in the \citet{Aigrain2015}
recovery experiment,
producing a large number of accurate rotation period measurements
(see, e.g., their Figure 8).
Another advantage is its fast implementation speed.
However, because the ACF method is non-probabilistic, ACF-estimated rotation
period uncertainties are poorly defined~---~a clear disadvantage.
It also requires evenly-spaced data, therefore is not applicable to
ground-based light curves.
Figure \ref{fig:demo_acf} contains an example ACF of the light curve in
Figure \ref{fig:demo_lc}.

We compare our results to those of the Tel Aviv team in \citet{Aigrain2015}
who use the ACF method described in \citet{Mcquillan2013}.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The injected and recovered rotation periods generally agree well: the MRD of
their periods, relative to the true rotation periods, is \percenttelavivMAD\%
--- smaller than the results produced by the GP method.
However, their results contain a larger number of extreme outliers --- they
predict periods that are significantly different from the true period in more
cases than the GP method.
This is revealed by comparing the RMS of the two sets of results since the RMS
is more sensitive to outliers.
The RMS of the Tel Aviv method results is \telavivRMS\ days and the RMS of GP
method results is \gpRMS\ days.
Another negative feature of the ACF method is revealed by examining the points
clustered close to the 1:1 line: more fall slightly below than above it, \ie\
rotation periods are systematically underestimated.
This stems from peak position measurements in the ACF method.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left (towards
shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice
simply measures the peak position without taking it into account.

@@

The \citet{Aigrain2015} Tel Aviv team were the only group to report periods
for all \naigrain\ stars, and are therefore the only team with which we
compare results.
The other participating groups chose to report only the period measurements in
which they were confident and are likely to have omitted large outliers.
In order therefore to test our results against another commonly used
technique, we implemented a simple LS periodogram method.

For each simulated light curve, we compute a LS periodogram\footnote{LS
periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.} over a
grid of 10,000 periods, evenly spaced in frequency, between 1 and 100 days.
We adopt the period of the highest peak in the periodogram as the measured
rotation period.
Figure \ref{fig:pgram_compare} shows the resulting recovered rotation periods
as a function of true period.
The MRD of the periodogram-recovered periods is \percentpgramMAD\%,
significantly worse than the ACF and GP methods (see table \ref{tab:MADs} for
a side-by-side comparison with the other methods).
The LS periodogram method drastically overestimates many rotation periods
because long-term trends unrelated to rotation can cause excess power at long
periods.

@@

\begin{table*}
\begin{center}
    \caption{Median absolute (MAD), median relative (MRD) deviations and
    root-mean-squared (RMS) for
    the LS periodogram, \citet{Mcquillan2013} ACF and GP period recovery
    methods.}
\begin{tabular}{lcccc}
    Method & MAD & MRD & RMS \\
    \hline
    GP (acf prior) & \gpMAD\ days & \percentgpMAD \% & \gpRMS\ days \\
    GP (uninformative prior) & \gpMADnp\ days & \percentgpMADnp \% &
    \gpRMSnp\ days \\
    ACF & \telavivMAD\ days & \percenttelavivMAD \%
    & \telavivRMS\ days \\
    LS periodogram & \pgramMAD\ days & \percentpgramMAD \% & \pgramRMS\ days\\
\end{tabular}
\end{center}
\end{table*}
\label{tab:MADs}

@@

Figure \ref{fig:mcquillan} compares the periods inferred with the GP method to
the ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping
KOIs.
This comparison shows generally very good agreement, demonstrating that this
method works not only on simulated data, but also on real data---with the
caveat that for any particular data set, some care is needed regarding the
setting the priors and ranges for the GP hyperparameters.
Notably, the GP method recovers periods
systematically slightly larger than \citet{Mcquillan2013}---a likely consequence
of correcting for the ACF peak measurement bias discussed in Section \ref{sec:acf}.
